---
title: "master_script_md"
author: "Savannah McNair"
date: "2025-04-12"
output: html_document
---

#1. Set up a local Spark server and add the two datasets you used in the first assignment from their GitHub repository. (20%)

#prerequisites
had to download java
```{r}
#system("java -version")
#install.packages("sparklyr")
#packageVersion("sparklyr")
library(sparklyr)
spark_install(version = "3.0")
```

#connecting - had to change the version to be the most recent compatible version
```{r}
sc <- spark_connect(master = "local", version = "3.0")
```


#2. In Spark, merge the two datasets, make a smaller version that includes only: Germany, China, Japan, United Kingdom, US, Brazil, Mexico and calculate the number of cases and rate of cases (cases/population) by country and day. Do two graphs and interpret them: change in the number of cases and change in rate by country. (25%)

#3. Run a ml_linear_regression explaining the log of number of cases using: country, population size and day since the start of the pandemic. Interpret the results. (25%)

#4. Write up everything in an analytic notebook (pdf Rmarkdown) that shows all the syntax you used, the results and walk the reader through the steps of your analysis. (20%)

#5. Presentation, presentation, presentation: README gives overview of the project and has session info, report text is easy to follow, graphs are easy to understand and properly formatted. (10%)
